---
title: "p8130_final"
author: "Zihan Lin"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(skimr)
library(naniar)
library(DataExplorer)
library(knitr)
library(kableExtra)
webshot::install_phantomjs()
library(ggplot2)
library(GGally)
library(janitor)
```

## R Markdown

```{r}
# Load the dataset
data <- read.csv("/Users/suwa/Desktop/p8130_final/data/Project_1_data.csv")

# Inspect the data structure
glimpse(data)  # Overview of the dataset

# Check for missing values
cat("Missing Values Summary:\n")
colSums(is.na(data))  # Count missing values per column

# Handle missing values
data <- data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# For categorical variables, impute missing with the mode
get_mode <- function(x) {
  unique_x <- unique(na.omit(x))
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

data <- data %>%
  mutate(across(where(is.character), ~ ifelse(is.na(.), get_mode(.), .)))

# Check for duplicates
cat("Checking for duplicate rows:\n")
sum(duplicated(data))  # Count duplicate rows

# Remove duplicates if any
data <- data %>% distinct()

# Check for invalid or inconsistent values
cat("Summary of score variables:\n")
summary(data$MathScore)
summary(data$ReadingScore)
summary(data$WritingScore)

# Replace invalid scores (e.g., >100 or <0) with NA
data <- data %>%
  mutate(
    MathScore = ifelse(MathScore < 0 | MathScore > 100, NA, MathScore),
    ReadingScore = ifelse(ReadingScore < 0 | ReadingScore > 100, NA, ReadingScore),
    WritingScore = ifelse(WritingScore < 0 | WritingScore > 100, NA, WritingScore)
  )

# Recheck missing values after cleaning
cat("Missing Values Summary After Cleaning:\n")
colSums(is.na(data))

# Recheck data types and convert if necessary
cat("Converting categorical variables to factors...\n")
data <- data %>%
  mutate(
    Gender = as.factor(Gender),
    EthnicGroup = as.factor(EthnicGroup),
    ParentEduc = as.factor(ParentEduc),
    LunchType = as.factor(LunchType),
    TestPrep = as.factor(TestPrep),
    ParentMaritalStatus = as.factor(ParentMaritalStatus),
    PracticeSport = as.factor(PracticeSport),
    IsFirstChild = as.factor(IsFirstChild),
    TransportMeans = as.factor(TransportMeans),
    WklyStudyHours = as.factor(WklyStudyHours)
  )

# Standardize numeric variables (if needed for modeling)
# Example: Scale test scores
data <- data %>%
  mutate(across(c(MathScore, ReadingScore, WritingScore), scale))

# Save Cleaned Data
write.csv(data, "/Users/suwa/Desktop/p8130_final/data/data_cleaned.csv", row.names = FALSE)
```

```{r}
# Reload the dataset
data <- read.csv("/Users/suwa/Desktop/p8130_final/data/data_cleaned.csv")

# Generate a summary table
skim(data)
```



```{r}
# Distributions of Test Scores
# Histograms for each test score
data %>%
  select(MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(everything(), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = Score, fill = Test)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "dodge") +
  labs(title = "Distributions of Test Scores", x = "Score", y = "Frequency") +
  theme_minimal()

# Boxplots for test scores
data %>%
  select(MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(everything(), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = Test, y = Score, fill = Test)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplots of Test Scores", x = "Test", y = "Score") +
  theme_minimal()

```

```{r}
# Boxplots for Categorical Covariates vs Test Scores
# Explore relationships between categorical covariates and test scores
data %>%
  pivot_longer(cols = c(MathScore, ReadingScore, WritingScore),
               names_to = "TestType", values_to = "Score") %>%
  ggplot(aes(x = Gender, y = Score, fill = Gender)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~TestType, scales = "free") +
  labs(title = "Test Scores by Gender",
       x = "Gender",
       y = "Score") +
  theme_minimal()
```

```{r}
# Distributions of Covariates
# Bar plots for categorical variables
categorical_vars <- data %>%
  select(where(is.factor))

for (var in names(categorical_vars)) {
  print(
    ggplot(data, aes_string(x = var, fill = var)) +
      geom_bar(alpha = 0.7) +
      labs(title = paste("Distribution of", var), x = var, y = "Count") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}

# Histograms for numeric covariates
numeric_vars <- data %>%
  select(where(is.numeric), -c(MathScore, ReadingScore, WritingScore))

for (var in names(numeric_vars)) {
  print(
    ggplot(data, aes_string(x = var)) +
      geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7) +
      labs(title = paste("Distribution of", var), x = var, y = "Frequency") +
      theme_minimal()
  )
}
```

```{r}
# Pairwise Relationships
# Scatterplots of test scores
data %>%
  ggplot(aes(x = MathScore, y = ReadingScore)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Scatterplot: Math vs. Reading Scores", x = "Math Score", y = "Reading Score") +
  theme_minimal()

data %>%
  ggplot(aes(x = MathScore, y = WritingScore)) +
  geom_point(alpha = 0.6, color = "green") +
  labs(title = "Scatterplot: Math vs. Writing Scores", x = "Math Score", y = "Writing Score") +
  theme_minimal()

data %>%
  ggplot(aes(x = ReadingScore, y = WritingScore)) +
  geom_point(alpha = 0.6, color = "purple") +
  labs(title = "Scatterplot: Reading vs. Writing Scores", x = "Reading Score", y = "Writing Score") +
  theme_minimal()
```


```{r}
# Correlation heatmap
corr_matrix <- cor(numeric_vars, use = "complete.obs")

corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "Variable1") %>%
  pivot_longer(cols = -Variable1, names_to = "Variable2", values_to = "Correlation") %>%
  ggplot(aes(x = Variable1, y = Variable2, fill = Correlation)) +
  geom_tile(color = "white") +  # Add grid lines
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmap", x = "Variable", y = "Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 10))
```

```{r}
# Boxplots for test scores by EthnicGroup
data %>%
  pivot_longer(cols = c(MathScore, ReadingScore, WritingScore), names_to = "TestType", values_to = "Score") %>%
  ggplot(aes(x = EthnicGroup, y = Score, fill = EthnicGroup)) +
  geom_boxplot() +
  facet_wrap(~ TestType, scales = "free") +
  theme_minimal() +
  labs(title = "Test Scores by Ethnic Group", x = "Ethnic Group", y = "Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Findings from Exploratory Data Analysis (EDA)

Pairwise Relationships:

Strong correlations between MathScore, ReadingScore, and WritingScore (r ≈ 0.95), suggesting redundancy in predictors for individual models. Weak correlation between NrSiblings and test scores. Visualizations indicate potential interaction effects, for example, between Gender and LunchType on MathScore.

Distributions:

Numeric variables like MathScore, ReadingScore, and WritingScore exhibit nearly normal distributions but with some skewness in scores below 50. NrSiblings is positively skewed with most values concentrated around 1 to 3.

Interactions and Covariate Effects:

Boxplots reveal that WklyStudyHours and EthnicGroup significantly impact test scores. Students with more than 10 hours of study time score higher across all test types.

Covariate Analysis:

Weekly study hours (WklyStudyHours) and test preparation (TestPrep) have clear separations in performance, suggesting strong predictive potential. Interaction plots highlight a differential impact of LunchType based on Gender.