---
title: "p8130_final"
author: "Zihan Lin"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(skimr)
library(caret)
library(naniar)
library(DataExplorer)
library(knitr)
library(kableExtra)
webshot::install_phantomjs()
library(ggplot2)
library(GGally)
library(janitor)
```

## R Markdown

```{r}
# Load the dataset
data <- read.csv("/Users/suwa/Desktop/p8130_final/data/Project_1_data.csv")

# Identify and encode binary categorical variables
# Check the structure of the data
str(data)

# Convert binary categorical variables to 0/1
binary_vars <- c("Gender", "LunchType", "TestPrep", "IsFirstChild", "TransportMeans")
data <- data %>%
  mutate(across(all_of(binary_vars), ~ ifelse(. == levels(as.factor(.))[1], 0, 1)))

# Create dummy variables for multi-category variables
# Identify multi-category variables
multi_category_vars <- c("EthnicGroup", "ParentEduc", "ParentMaritalStatus", "PracticeSport", "WklyStudyHours")

# Generate dummy variables for multi-category variables
dummy_vars <- dummyVars("~ .", data = data, fullRank = TRUE)
data <- predict(dummy_vars, newdata = data) %>% as.data.frame()

# Handle missing values
# Remove rows with missing values
data <- data %>% drop_na()

# Check the cleaned dataset
str(data)        # Check the structure of the cleaned dataset
summary(data)    # Summarize the cleaned data
head(data)       # View the first few rows of the cleaned data

# Save Cleaned Data
write.csv(data, "/Users/suwa/Desktop/p8130_final/data/data_cleaned.csv", row.names = FALSE)
```

```{r}
# Reload the dataset
data <- read.csv("/Users/suwa/Desktop/p8130_final/data/data_cleaned.csv")

# Generate a summary table
skim(data)
```



```{r}
# Distributions of Test Scores
# Histograms for each test score
data %>%
  select(MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(everything(), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = Score, fill = Test)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "dodge") +
  labs(title = "Distributions of Test Scores", x = "Score", y = "Frequency") +
  theme_minimal()

# Boxplots for test scores
data %>%
  select(MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(everything(), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = Test, y = Score, fill = Test)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplots of Test Scores", x = "Test", y = "Score") +
  theme_minimal()

```

```{r}
# Distributions of Categorical Covariates
# Identify original categorical variables in the data
categorical_vars <- c("Gender", "LunchType", "TestPrep", "IsFirstChild")

# Bar plots for categorical variables
for (var in categorical_vars) {
  print(
    ggplot(data, aes_string(x = var, fill = var)) +
      geom_bar(alpha = 0.7) +
      labs(title = paste("Distribution of", var), x = var, y = "Count") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r}
# Relationships Between Test Scores and Key Covariates
# Boxplots for test scores by Gender
data %>%
  select(Gender, MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(cols = starts_with("MathScore"), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = Gender, y = Score, fill = Gender)) +
  geom_boxplot() +
  facet_wrap(~ Test) +
  labs(title = "Test Scores by Gender", x = "Gender", y = "Score") +
  theme_minimal()

# Boxplots for test scores by LunchType
data %>%
  select(LunchType, MathScore, ReadingScore, WritingScore) %>%
  pivot_longer(cols = starts_with("MathScore"), names_to = "Test", values_to = "Score") %>%
  ggplot(aes(x = LunchType, y = Score, fill = LunchType)) +
  geom_boxplot() +
  facet_wrap(~ Test) +
  labs(title = "Test Scores by Lunch Type", x = "Lunch Type", y = "Score") +
  theme_minimal()
```

```{r}
# Pairwise Relationships
# Scatterplots of test scores
data %>%
  ggplot(aes(x = MathScore, y = ReadingScore)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Scatterplot: Math vs. Reading Scores", x = "Math Score", y = "Reading Score") +
  theme_minimal()

data %>%
  ggplot(aes(x = MathScore, y = WritingScore)) +
  geom_point(alpha = 0.6, color = "green") +
  labs(title = "Scatterplot: Math vs. Writing Scores", x = "Math Score", y = "Writing Score") +
  theme_minimal()

data %>%
  ggplot(aes(x = ReadingScore, y = WritingScore)) +
  geom_point(alpha = 0.6, color = "purple") +
  labs(title = "Scatterplot: Reading vs. Writing Scores", x = "Reading Score", y = "Writing Score") +
  theme_minimal()
```


```{r}
# Correlation heatmap
numeric_vars <- data %>% select(where(is.numeric))
corr_matrix <- cor(numeric_vars, use = "complete.obs")

corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "Variable1") %>%
  pivot_longer(cols = -Variable1, names_to = "Variable2", values_to = "Correlation") %>%
  ggplot(aes(x = Variable1, y = Variable2, fill = Correlation)) +
  geom_tile(color = "white") +  # Add grid lines
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmap", x = "Variable", y = "Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 10))
```


### Findings from Exploratory Data Analysis (EDA)

Pairwise Relationships:

Strong correlations between MathScore, ReadingScore, and WritingScore (r ≈ 0.95), suggesting redundancy in predictors for individual models. Weak correlation between NrSiblings and test scores. Visualizations indicate potential interaction effects, for example, between Gender and LunchType on MathScore.

Distributions:

Numeric variables like MathScore, ReadingScore, and WritingScore exhibit nearly normal distributions but with some skewness in scores below 50. NrSiblings is positively skewed with most values concentrated around 1 to 3.

Interactions and Covariate Effects:

Boxplots reveal that WklyStudyHours and EthnicGroup significantly impact test scores. Students with more than 10 hours of study time score higher across all test types.

Covariate Analysis:

Weekly study hours (WklyStudyHours) and test preparation (TestPrep) have clear separations in performance, suggesting strong predictive potential. Interaction plots highlight a differential impact of LunchType based on Gender.